**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**

### 2021-01-20 Manuscripts/Manuscript_AM_v2.Rmd
Updated

## The Basis function approach

@@

Where $y_{ijt}$ is the response at time $t$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta)$ and $\epsilon_{ijt}$ represents the residuals.

In contrast to rm-ANOVA or LMEMs, GAMs use _smooth functions_  to model the relationship between the covariates and the response. This approach is more advantageous as it does not restrict the model to a linear relationship. One possible function for $f(x_t)$  is a polynomial, but a major limitation is that they create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].

To overcome this limitation, the smooth functions in GAMs  are represented using _basis functions_  over  evenly spaced ranges of the covariates known as _knots_. The _basis function_ used is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together[@wood2017;@simpson2018]. Cubic splines have a long history in their use to solve non-parametric statistical problems and are the default choice to fit GAMs as they are the simplest option to obtain visual smoothness [@wegman1983]. Therefore, GAMs overcome the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear, such as \@ref(fig:quadratic-response). Regarding longitudinal data, Pedersen et al [@pedersen2019] demonstrated the capabilities of GAMs in this area using ecological data.

>>>>

## GAMs and Basis Functions

@@

Where $y_{ijt}$ is the response at time $t$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta)$ and $\varepsilon_{ijt}$ represents the residuals.

In contrast to rm-ANOVA or LMEMs, GAMs use _smooth functions_  to model the relationship between the covariates and the response. This approach is more advantageous as it does not restrict the model to a linear relationship. One possible function for $f(x_t\mid \beta)$ that allows for non-linear responses is a polynomial, but a major limitation is that they create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].In particular, as $t$ goes to $\pm \infty$, $f(x_t \mid \beta)$ goes to $\pm \infty$ which is almost always unrealistic.

The model specification for GAMs requires that the _smooth functions_ are represented in a parametric way that fit within the GLM framework, and this step is achieved by using a _basis function_ to represent such functions. A _basis_ is a set of functions that capture the space where the smooths that approximate $f(x_t \mid \beta)$ exist [@simpson2018]. In other words, 
To overcome this limitation, the smooth functions in GAMs  are represented using _basis functions_  over  evenly spaced ranges of the covariates known as _knots_. The _basis function_ used is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together[@wood2017;@simpson2018]. Cubic splines have a long history in their use to solve non-parametric statistical problems and are the default choice to fit GAMs as they are the simplest option to obtain visual smoothness [@wegman1983]. Therefore, GAMs overcome the limitation that occurs in LMEMs and rm-ANOVA when the data is non linear, such as \@ref(fig:quadratic-response). Regarding longitudinal data, Pedersen et al [@pedersen2019] demonstrated the capabilities of GAMs in this area using ecological data.
